### 2025-09-22 pre_print/response_to_reviewers.qmd

---
title: "Response to Reviewers"
subtitle: "The effects of supervision on effort during resistance training: A Bayesian analysis of prior data and an experimental study of private strength clinic members"
format: pdf
bibliography: bibliography.bib  
---

@@

Dear Professor Gruet,

Many thanks for considering our manuscript and placing it under further review with *Peer Community in Health and Movement Sciences*.

We apologise sincerely for our delay in addressing these and resubmitting. Unfortunately other commitments and workloads had prevented us from returning to this. We have attempted to address or rebut the additional comments provided by the reviewers.

@@

### Author Response

We have made this change.

### Reviewer Comment

Abstract - TUL, RPE, and RPD are measures of effort, not exercise performance. There are several instances in which “exercise performance” is used in place of “effort,” wherein only the latter should be used. Be consistent with the constructs and operational definitions.

@@

### Author Response

In the present study exercise performance as TUL is used as an operationalisation to draw inferences about effort. As such, in places where we are drawing broader inferences about the concept of effort as informed by the exercise performance we use "effort", but in others we use more accurately to the operationalisation "exercise performance" or "TUL". We have added the following the abstract to clarify this:

*"...exercise performance, measured as time under load (TUL) where a longer TUL at a given load is indicative of greater effort..."*

We have also added to the introduction the text in parentheses at the end of this sentence to clarify this conceptually:

*"Effort is conceptualised at the relation of task demands to the capacity to meet those demands [@steeleWhatPerceptionEffort2020] and so in RT is determined by both the load utilised and also the proximity to momentary failure due to the fatigue (i.e., reduction in capacity) that occurs with continued performance at a given load (i.e., with more repetitions or a longer time under load)."*

And clarified in the methods:

*"The TUL performed (where a longer TUL was interpreted as indicating greater effort), and ratings of perceptions of effort (RPE) and discomfort (RPD) were recorded and compared between both conditions."*

@@

### Reviewer Comment

Abstract - Second sentence: “We investigated supervision’s role in effort…” – “in” should be “on”

@@

### Author Response

We have made this change.

### Reviewer Comment

The Introduction describes the existing knowledge of the effect of supervision on resistance training outcomes, specifically strength, and summarizes the study’s design and methods. While the essential content is present, it could be more clearly discussed. I suggest reorganizing the introduction as follows: 
• The first paragraph is good as is. 
• I would present the hypothesized mechanisms together up front, i.e., trainers enforce load progression and help motivate the exercisers to apply the proper amount of effort (i.e., closer proximity to momentary muscular failure). 
• I would then discuss the evidence for each hypothesized mechanism, which is mostly indirect. 
• The second-to-last paragraph should summarize the existing gaps in knowledge, i.e., the data concerning supervision effects are still limited, especially in ecological settings, and the potential mechanisms by which supervision might enhance strength are likewise unclear.  
• The final paragraph should more explicitly state the study purpose and along with a hypothesis. For example, “The purpose of this study was to examine the effects of supervision on client effort in resistance training in a more ecological setting. We hypothesized that … I would then introduce the methods as the authors have done by stating that “We tested the hypothesis by …”

@@

### Author Response

We thank the reviewer for their suggested revision to the introduction. We have however opted not to make any major rewrites to it as we do feel the major rationale is presented clearly already, and we have not added any hypothesis as our analyses were not aimed at any kind of statistical hypothesis testing but instead exploratory and descriptive parameter estimation. Lastly, given the suggested restructure is largely one of preference, this is somewhat of a pragmatic choice on our part to not make changes because of available time for making revisions given other workload demands currently (and that we're already pretty late with making revisions in response to this second round of review - our apologies!).

@@

### Reviewer Comment

Minor comments on the writing: There were several errors and suboptimal wording choices present. 
• The word “recent” (typically followed by “meta-analysis”) is repetitively used throughout. Just state the result and cite the paper. For example, the second sentence, “In a recent systematic review and meta-analysis from Fisher et al. (2022) there was a moderate standardised mean effect (0.40 [95%CI: 0.06, 0.74]) of supervised vs unsupervised RT on strength outcomes synthesised from ten studies” could be more succinctly stated as “A meta-analysis of 10 studies demonstrated a moderate benefit of supervision on strength outcomes (standardized mean effect = 0.40 [95%CI: 0.06, 0.74]; Fisher et al., 2022).” 
• Error, second paragraph: “Effort is conceptualized at…” – “at” should be “as” 
• Grammar, third paragraph: “But a caveat is that there was limited data…” – the word “data” is plural, so “was” should be “were.” There are other instances of this error in the manuscript.

@@

### Author Response

We have however made these minor changes as noted for the existing introduction.

@@

### Reviewer Comment

As with the Introduction, the Methods are thoroughly described but could be better organized to highlight critical features. In particular, the operational definitions of effort in this study were TUL, RPE, and RPD. These operational definitions should be more explicitly defined and justified as measures of effort. Currently, their description is an afterthought buried in a paragraph: “The TUL performed, and ratings of perceptions of effort (RPE) and discomfort (RPD) were recorded and compared between both conditions.” Furthermore, whether session RPE or RPE post each exercise is not made clear until the end of the first paragraph in the Protocols section (“In addition, they recorded their RPE and RPD in that order immediately upon completing the exercise using previously validated scales for differentiating these perceptions.”) Such critical details about the measures should be mentioned upfront in their own paragraph.

@@

### Author Response

As above with the introduction revisions, we have opted not to make further structural changes as we believe that the required information is contained in the methods section. As noted above though we have more clearly noted that, given the conceptualisation of effort, the exercise performance as TUL is indicative as an operationalisation of effort in the present study.

@@

### Reviewer Comment

I question the following sentence: “This study was not pre-registered and as described below the sample size was justified based on logistical concerns and the analysis is considered exploratory.” To me, it seems a bit disingenuous to state the analysis is exploratory when there are meta-analyses of supervision effects on RT outcomes and a clear testable hypothesis that the study addresses using a very detailed Bayesian analysis. Why was the experimental intervention not pre-registered? Was this a deliberate choice? Could the study be retrospectively registered? (I have seen others do so.)

@@

### Author Response

We have tried to be as transparent as possible with this study. The original experimental study was conducted several years previously, and before any of the authors were fully cognisant of the value of pre-registration. Further, we did not have a sufficiently precise statistical hypothesis to test for which we could for example have performed a power analysis (Frequentist or Bayesian) to help justify sample size for the experimental hypothesis. We would prefer not to retrospectively register given this would be merely performative at this stage and does not serve the primary purpose of pre-registration (i.e., the transparent reporting of research to enable the third-party evaluation of the severity of tests for claims made based on hypothesis tests). In retrospect we could have looked to convert the standardised effect size metrics to the relevant scales for the parameter estimates presented contrasting between conditions, and then conduct either an analysis akin to examining the overlap with the region of practical equivalence (ROPE), or calculated Bayes Factors for this. But we feel this would be disingenuous to add (and also take more time than we have currently) as we never planned to do this a priori. As such, having the full dataset open and being transparent about the nature of the study is appropriate in our opinion (notably, these responses will also be publicly available so people can see why we made this choice).

@@

### Reviewer Comment

Experimental approach section: This sentence is run-on: “This consists of a single set of the resistance machine exercises prescribed on their current training programme card using a load that should permit them to reach momentary failure within a time-under-load (TUL) of 90-120 seconds (though an upper limit of 180 seconds TUL is enforced to avoid machines being occupied for too long on the clinic floor preventing other members from using them) using a ~ 12 seconds repetition duration (i.e., ~4:4 seconds concentric:eccentric actions with a 2 isometric second hold whilst still under load with tension, not “locked-out”, on the involved musculature at the end of each concentric and eccentric muscle action).”

@@

### Author Response

We have split this sentence now.

*"This consists of a single set of the resistance machine exercises prescribed on their current training programme card. A load is used that should permit them to reach momentary failure within a time-under-load (TUL) of 90-120 seconds (though an upper limit of 180 seconds TUL is enforced to avoid machines being occupied for too long on the clinic floor preventing other members from using them) using a \~ 12 seconds repetition duration (i.e., \~4:4 seconds concentric:eccentric actions with a 2 isometric second hold whilst still under load with tension, not "locked-out", on the involved musculature at the end of each concentric and eccentric muscle action)."*

@@

### Reviewer Comment

Previous observational sample subsection: The following sentence is run on, contains several awkward passages, and contains punctuation errors: “For TUL we limited data to Core and Assisted members training sessions that either were not led by an exercise scientist or were respectively, took the first training session after at least 6 months of previous training at Kieser Australia had been completed by each member, randomly sampled 1000 Core and 1000 Assisted members and then filtered to the resistance machines used in by members in the experimental study so that we had a selection of members across varied clinic locations and completing sessions with a selection of resistance machine exercises and had TUL data for each exercise.” Suggest rewriting as a numbered list of inclusion criteria for the TUL data.

@@

### Author Response

We have rewritten this as:

*"For TUL we limited data to Core and Assisted members training sessions that either were not, or were, led by an exercise scientist respectively[^3]. We took the first training session after at least 6 months of previous training at Kieser Australia had been completed by each member. Then we randomly sampled 1000 Core and 1000 Assisted members and filtered to the resistance machines used by members in the experimental study."*

@@

### Reviewer Comment

Protocols section:  
• The first paragraph is overlong – a new paragraph could begin at the “Participants were instructed to ensure…”  
• Typo: “These TUL recordings where then input…” – “where” should be “were”

@@

### Author Response

We have made both of these changes.

### Reviewer Comment

Statistical Analysis:  
• “All posterior estimates and their precision,…” Precision should be pluralized as “precisions” 
• Paragraph starting with “For both the previous observational sample…” The authors define “session type” (core or assisted) but then refer to it as “condition” in the model description. Suggest helping the reader by using consistent terminology.

@@

### Author Response

We have pluralised "precisions" and also added *"...session type (i.e., condition)..."* to clarify the terminology.

Figure 2 title: “Prior Sample Data Distributions” should be changed to “Previous observational sample histograms”

We have changed the title.

An issue I sensed with the first version of the paper but failed to articulate in my original comments was the lack of validation of the modeling approach. The study features many assumptions and modeling decisions, and the reader is left to wonder about the extent to which the conclusions are robust to them. A comment on the sensitivity of the conclusions to the modeling decisions would be appropriate in the Discussion.

Whilst we agree to some extent that all elements of model selection and the assumptions underlying them should be well justified and explained the standards are often double in this regard - researchers in our field using more traditional t-test or ANOVA type analyses (and which probably don't even understand the assumptions underlying them) don't get asked to justify these. We've explained quite clearly, and also provided the lay summary for our modelling choices. Adding a comment to the discussion about the possible sensitivity of our conclusions to our choice of models is in one sense tautological as our results, as are any using a single modelling approach, conditional on that model (the exception being in the case of Bayesian Model Averaging) but also that making that point leads to the question why not take "multiverse" type approach and try out lots of different models to see what happens. In principle this would be great, but as we've noted we don't have time to continue working on this in that fashion. If readers are curious then the data is openly available and they are welcome to examine whether a different, perhaps more justifiable, modelling choice would lead to substantively different estimates.

In addition, diagnostic analyses were performed and included as a supplementary file, which I commend, but they were not mentioned in the Results or Discussion. A quick sentence or two summarizing the diagnostics would help assure the reader about the model validities.

We have added to the analysis section where we link to the diagnostic plots:

*"Trace plots were produced along with $\hat{R}$ values to examine whether chains had converged (all were $\hat{R}<1.05$ indicating convergence)..."*

As I alluded to above, the study purpose could be more explicitly articulated in the first paragraph of the Discussion. I suggest modifying the first sentence as follows: “Our study is a first to offer insights into the role of supervision on exerciser effort during RT in an ecologically valid real-world setting. These insights were derived from a unique and strong study design involving observational and experimental components.”

I suggest modifying the second sentence as follows to focus on the outcomes measures as operational definitions of effort: “…we assessed effort as the TUL, RPE, and RPD.”

We have changed these sentence to the ones suggested.

Some paragraphs are overlong. The limitations paragraph, for example, takes up a whole page. It could be broken up into single paragraphs for each of the ~4 limitations discussed by the authors.

We have broken up the limitations paragraph.

@@

We have broken up the limitations paragraph.

## Reviewer 2 - Anonymous

### Reviewer Comment

I encourage the authors to revise the conclusion so that it: 
1. Maintains the statement that under real-world settings there was little difference in exercise performance. 
2. Clearly states that the observed benefit in the experimental setup was based on an acute effect of a single session. 
3. Avoids speculation and focuses on direct interpretations of the data.

@@

### Author Response

We have made changes to the conclusions to try to more accurately reflect the findings as suggested (and also minor change to the abstract). The conclusion now reads:

*"In real world settings there was little difference in exercise performance suggesting similar degrees of effort during both supervised and unsupervised conditions. Yet, in our experimental study there was a clear benefit to performance when under supervision suggesting trainees performed with a higher degree of effort. There was a small difference in RPE reported with and without supervision under both real world and experimental conditions suggesting that under supervision trainees train with greater proximity to failure, which was also supported by greater RPD under supervision; though in real world conditions the RPE reported suggested that effort may be sub-optimal irrespective of condition. These results in general support previous work highlighting the importance of supervision during RT and that under experimental conditions trainees likely train with greater effort when supervised."*

### 2025-09-22 pre_print/pre_print_v3.qmd
Second round of manuscript revisions

---
articletitle: |
  The effects of supervision upon effort during resistance training: A Bayesian analysis of previous observational data and an experimental study of private strength clinic members
# shorttitle: "Doe et al. (2022)" # usually auto-generated
format: 
  sportrxiv-pdf:
    include-in-header:
      text: |
       \usepackage[font=scriptsize]{caption}
       \usepackage{caption}
author:
  - name: James Steele
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
  - name: James Fisher
    affiliations:
      - ref: 3
      - ref: 4
    orcid: 0000-0002-6013-8402
    email: james.fisher.phd@outlook.com
  - name: Tim Dettmann
    affiliations:
      - ref: 1
    email: tim.dettmann@gmail.com
affiliations:
      - id: 1
        name: Kieser Australia, Melbourne, Australia
      - id: 2
        name: Steele Research Limited, Eastleigh, UK
      - id: 3
        name: Department of Sport and Health, Solent University, Southampton, UK
      - id: 4
        name: The Exercise Coach, Tennessee, USA
abstract: |
  Supervision during resistance training (RT) may enhance strength gains by optimizing trainee effort. We investigated supervision’s role in effort during RT in a unique setting with private strength clinics, where members train either unsupervised (“Core” membership) or supervised by a qualified exercise scientist (“Assisted” membership). Using both retrospective analysis of member training records and a prospective experimental study, we examined supervision’s impact on exercise performance, measured as time under load (TUL), rating of perceived effort (RPE), and rating of perceived discomfort (RPD). Bayesian methods were applied, using empirically informed prior distributions from retrospective data to model the experimental study. The previous observational sample included ~1000 members training sessions from each membership type, while the experimental study involved 45 Core members performing both supervised and unsupervised sessions in randomized order, using their current training loads to momentary failure. Our findings suggest that, in real-world settings (in situ), exercise performance differed little between supervised and unsupervised training. However, in our experimental study, supervision improved TUL (Core = 125.12 [95%QI: 113.70, 131.90] sec; Assisted = 147.35 [95%QI: 134.29, 154.81] sec; contrast = -22.10 [95%QI: -26.60, -17.61] sec). In percentage points RPE was slightly higher with supervision in both previous observational real-world (Core = 53% [95%QI: 51%, 55%]; Assisted = 59% [95%QI: 57%, 61%]; contrast = -6% [95%QI: -8%, -4%]) and experimental settings (Core = 81% [95%QI: 75%, 86%]; Assisted = 87% [95%QI: 83%, 91%]; contrast = -6% [95%QI: -10%, -4%]), suggesting trainees push closer to failure under supervision. This was further supported by higher RPD during the experimental study (Core = 6.3 [95%QI: 5.1, 7.3]; Assisted = 7.5 [95%QI: 6.5, 8.3]; contrast = -1.2 [95%QI: -1.6, -0.9]). Overall, these results reinforce research on the benefits of supervision in RT, indicating that unsupervised trainees—especially in real-world conditions—likely train with suboptimal effort.
license-type: ccby # change if neccessary
# year: 2025 # defaults to current year
keywords: [supervision, effort, resistance training, bayesian] # optional
# optional link to Supplementary Material:
suppl-link: https://osf.io/etb34/
reference-section-title: References
printnote: "PREPRINT - NOT PEER REVIEWED" # if post print, include "POSTPRINT" then link to the published article
bibliography: bibliography.bib  
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

Recent work has highlighted the importance of supervision in optimising strength outcomes from resistance training (RT). In a recent systematic review and meta-analysis from Fisher et al. [-@fisherRoleSupervisionResistance2022] there was a moderate standardised mean effect (0.40 \[95%CI: 0.06, 0.74\]) of supervised vs unsupervised RT on strength outcomes synthesised from ten studies. Since then, a further study has reported that even in previously trained participants there may be benefit to strength outcomes from supervision [@colemanSupervisionResistanceTraining2023].

It has been speculated [@fisherRoleSupervisionResistance2022] that an explanation for this effect may be due to the role that a trainer plays in prescribing load progression for the trainee, whereas when there is no clear load progression rule unsupervised trainees may be less likely to train with appropriate loads, and ultimately appropriate effort. Effort is conceptualised at the relation of task demands to the capacity to meet those demands [@steeleWhatPerceptionEffort2020] and so in RT is determined by both the load utilised and also the proximity to momentary failure due to the fatigue (i.e., reduction in capacity) that occurs with continued performance.

A recent meta-analysis examining the relative loads selected by trainees when given the ability to self-select highlights that they tend to choose loads which, whilst initially efficacious in novice trainees, become sub-optimal quickly as training experience progresses (\~53% of one repetition maximum \[1RM\]) particularly when combined with the typical repetition ranges prescribed i.e., \~8 to 15 repetitions [@steeleAreTraineesLifting2022]. Further, when given the opportunity to self-select load and the number of repetitions to complete there is evidence that trainees likely train with relatively low effort as determined from their proximity to momentary failure i.e., they select \~10 repetitions at \~53% 1RM [@steeleAreTraineesLifting2022; @nuzzoMaximalNumberRepetitions2024]. Thus, whether self-selecting a load for a typically prescribed repetition range or self-selecting a load and self-selecting the number of repetitions to perform, most will naturally perform sets with an estimated proximity of \~10-20 repetitions shy of momentary failure.

@@

However, a recent meta-analysis of the dose-response relationship of proximity to momentary failure for strength and hypertrophy outcomes highlights that, while there is increased hypertrophy with closer proximity to failure, there is not a clear relationship for strength [@robinsonExploringDoseResponse2024]. But a caveat is that there was limited data for proximities to failure \>10 repetitions, and that the models were adjusted for load (average loads were typically \~75-85% of one repetition maximum). Thus, the results of this meta-analysis apply to proximity to failure *after* selecting load (i.e., intraset effort) and a higher relative load requires a greater effort all else being equal. When considering the average self-selected load of \~53% 1RM coupled with a typical \~10 repetitions per set equating to proximities to failure \>10 repetitions [@steeleAreTraineesLifting2022] it seems plausible that unsupervised trainees may train with a less-than-optimal load, repetitions, and resultant effort to optimise strength outcomes. Ultimately, and in combination, supervision should provide trainees with feedback of their past performance, whether technical, effort- or program-based, and guidance toward their future performance [@fisherSupervisionStrengthTrainingthe2025].

@@

The study was conducted at a selection of strength clinics operated by Kieser Australia recruiting from the existing pool of Core members at these locations. All Kieser Australia members are prescribed the same protocol. This consists of a single set of the resistance machine exercises prescribed on their current training programme card using a load that should permit them to reach momentary failure within a time-under-load (TUL) of 90-120 seconds (though an upper limit of 180 seconds TUL is enforced to avoid machines being occupied for too long on the clinic floor preventing other members from using them) using a \~ 12 seconds repetition duration (i.e., \~4:4 seconds concentric:eccentric actions with a 2 isometric second hold whilst still under load with tension, not "locked-out", on the involved musculature at the end of each concentric and eccentric muscle action). Core trainees are prompted to progress load by \~5% for the next session once a TUL of \>120 seconds is achieved for a particular exercise before reaching momentary failure, and for Assisted trainees this is actioned by their supervising Exercise Scientist. Trainees are prompted to complete a session rating of perceived effort (RPE) at the end of their training session.

An acute randomised cross-over design was employed to examine participants completing an RT session with and without direct supervision with one week apart (i.e., a Core vs Assisted session). Participants were instructed to complete a single set of the resistance machine exercises on their current training programme card using their current training loads to momentary failure. The TUL performed, and ratings of perceptions of effort (RPE) and discomfort (RPD) were recorded and compared between both conditions. This study was not pre-registered and as described below the sample size was justified based on logistical concerns and the analysis is considered exploratory.

@@

We queried the Kieser Australia database to generate random samples from historical data. We opted to generate reasonably large samples to ensure precise prior estimates, though did not use the entirety of the historical data (\~50000 members, \>10 years of data) so as to reduce computation time for Bayesian modelling. For TUL we limited data to Core and Assisted members training sessions that either were not led by an exercise scientist or were respectively[^3], took the first training session after at least 6 months of previous training at Kieser Australia had been completed by each member, randomly sampled 1000 Core and 1000 Assisted members and then filtered to the resistance machines used in by members in the experimental study so that we had a selection of members across varied clinic locations and completing sessions with a selection of resistance machine exercises and had TUL data for each exercise. Thus our final sample size was n = `r length(unique(prior_data_tul$id))` (Core = `r length(unique(filter(prior_data_tul, core_assisted == "core")$id))`, Assisted = `r length(unique(filter(prior_data_tul, core_assisted == "assisted")$id))`). For session RPE we performed a similar database query and randomly sampled 1000 Core and 1000 Assisted members who had reported session RPE values.

@@

All RT was performed using resistance machines (Kieser Training AG, Zurich, Switzerland) including the A1 (hip extension), A2 (torso flexion), A3 (hip abduction), B1 (leg extension), B6 (leg press), B7 (seated leg curl), C3 (torso arm i.e., pulldown), C7 (seated row), D5 (arm cross i.e., pectoral fly), D6 (chest press), D7 (seated dip), F2/F2.1 (abdominal flexion), and K2 (supported supinated grip pullup). Note, not all participants performed exercise using each machine, but had at least some of these machines in their current training cards. For the experimental study participants were instructed to attend two training sessions at least one-week apart where they completed either their current Core session (i.e., unsupervised) or an Assisted session (i.e., supervised). During the Core session the participant completed their training session as prescribed with the only difference for the Assisted session being that an exercise scientist supervised them providing instruction and motivation; however, they did not interfere in any physical way during the set (e.g. to spot or assist in completing a failing repetition). Each participant had an existing training programme card. In both conditions they utilised the current training loads for each machine that were recorded in the last session of their training programme card before participation in the study. Participants were instructed to ensure that they continue performing the exercise to momentary failure independently of the TUL achieved for the purposes of the study i.e., if they realised during the set that their previous load selection was evidently too low for the prescribed TUL range they should continue to momentary failure regardless off the TUL acheived. Momentary failure was defined as per Steele et al. [-@steeleClarityReportingTerminology2017] i.e., the point at which, despite their greatest effort, participants are unable to continue concentrically contracting and moving the resistance, and this was clearly communicated to the participants. During each session for each exercise the trainees/trainers recorded their TUL achieved as they would do for their usual sessions using standard timers situated around the clinic in view during training specifically for this purpose (these are essentially Swiss engineered clocks with sweep movement for the second hand and positioned so that there is always one in eye line for every machine to enable recording of the TUL in seconds). These TUL recordings where then input to their training cards. In addition, they recorded their RPE and RPD in that order immediately upon completing the exercise using previously validated scales for differentiating these perceptions [@steeleDifferentiationPerceivedEffort2017]; scripts and scales are available here: [https://osf.io/ufvy8/](https://osf.io/ufvy8/).

>>>>

---
articletitle: |
  The effects of supervision on effort during resistance training: A Bayesian analysis of previous observational data and an experimental study of private strength clinic members
# shorttitle: "Doe et al. (2022)" # usually auto-generated
format: 
  sportrxiv-pdf:
    include-in-header:
      text: |
       \usepackage[font=scriptsize]{caption}
       \usepackage{caption}
author:
  - name: James Steele
    affiliations:
      - ref: 1
      - ref: 2
      - ref: 3
    orcid: 0000-0002-8003-0757
    corresponding: true
    email: james@steele-research.com
  - name: James Fisher
    affiliations:
      - ref: 3
      - ref: 4
    orcid: 0000-0002-6013-8402
    email: james.fisher.phd@outlook.com
  - name: Tim Dettmann
    affiliations:
      - ref: 1
    email: tim.dettmann@gmail.com
affiliations:
      - id: 1
        name: Kieser Australia, Melbourne, Australia
      - id: 2
        name: Steele Research Limited, Eastleigh, UK
      - id: 3
        name: Department of Sport and Health, Solent University, Southampton, UK
      - id: 4
        name: The Exercise Coach, Tennessee, USA
abstract: |
  Supervision during resistance training (RT) may enhance strength gains by optimizing trainee effort. We investigated supervision’s role on effort during RT in a unique setting with private strength clinics, where members train either unsupervised (“Core” membership) or supervised by a qualified exercise scientist (“Assisted” membership). Using both retrospective analysis of member training records and a prospective experimental study, we examined supervision’s impact on exercise performance, measured as time under load (TUL) where a longer TUL at a given load is indicative of greater effort, rating of perceived effort (RPE), and rating of perceived discomfort (RPD). Bayesian methods were applied, using empirically informed prior distributions from retrospective data to model the experimental study. The previous observational sample included ~1000 members training sessions from each membership type, while the experimental study involved 45 Core members performing both supervised and unsupervised sessions in randomized order, using their current training loads to momentary failure. Our findings suggest that, in real-world settings (in situ), exercise performance differed little between supervised and unsupervised training with trainees likely using suboptimal effort under both conditions. However, in our experimental study, supervision improved TUL (Core = 125.12 [95%QI: 113.70, 131.90] sec; Assisted = 147.35 [95%QI: 134.29, 154.81] sec; contrast = -22.10 [95%QI: -26.60, -17.61] sec). In percentage points RPE was slightly higher with supervision in both previous observational real-world (Core = 53% [95%QI: 51%, 55%]; Assisted = 59% [95%QI: 57%, 61%]; contrast = -6% [95%QI: -8%, -4%]) and experimental settings (Core = 81% [95%QI: 75%, 86%]; Assisted = 87% [95%QI: 83%, 91%]; contrast = -6% [95%QI: -10%, -4%]), suggesting trainees push closer to failure under supervision. This was further supported by higher RPD during the experimental study (Core = 6.3 [95%QI: 5.1, 7.3]; Assisted = 7.5 [95%QI: 6.5, 8.3]; contrast = -1.2 [95%QI: -1.6, -0.9]). Overall, these results reinforce research on the benefits of supervision in RT, indicating that under experimental conditions trainees likely train with greater effort when supervised.
license-type: ccby # change if neccessary
# year: 2025 # defaults to current year
keywords: [supervision, effort, resistance training, bayesian] # optional
# optional link to Supplementary Material:
suppl-link: https://osf.io/etb34/
reference-section-title: References
printnote: "PREPRINT - NOT PEER REVIEWED" # if post print, include "POSTPRINT" then link to the published article
bibliography: bibliography.bib  
pdf-engine: xelatex
execute: 
  echo: false
  message: false
  warning: false
---

@@

Recent work has highlighted the importance of supervision in optimising strength outcomes from resistance training (RT). In a systematic review and meta-analysis from Fisher et al. [-@fisherRoleSupervisionResistance2022] there was a moderate standardised mean effect (0.40 \[95%CI: 0.06, 0.74\]) of supervised vs unsupervised RT on strength outcomes synthesised from ten studies. Since then, a further study has reported that even in previously trained participants there may be benefit to strength outcomes from supervision [@colemanSupervisionResistanceTraining2023].

It has been speculated [@fisherRoleSupervisionResistance2022] that an explanation for this effect may be due to the role that a trainer plays in prescribing load progression for the trainee, whereas when there is no clear load progression rule unsupervised trainees may be less likely to train with appropriate loads, and ultimately appropriate effort. Effort is conceptualised as the relation of task demands to the capacity to meet those demands [@steeleWhatPerceptionEffort2020] and so in RT is determined by both the load utilised and also the proximity to momentary failure due to the fatigue (i.e., reduction in capacity) that occurs with continued performance at a given load (i.e., with more repetitions or a longer time under load)).

A meta-analysis examining the relative loads selected by trainees when given the ability to self-select highlights that they tend to choose loads which, whilst initially efficacious in novice trainees, become sub-optimal quickly as training experience progresses (\~53% of one repetition maximum \[1RM\]) particularly when combined with the typical repetition ranges prescribed i.e., \~8 to 15 repetitions [@steeleAreTraineesLifting2022]. Further, when given the opportunity to self-select load and the number of repetitions to complete there is evidence that trainees likely train with relatively low effort as determined from their proximity to momentary failure i.e., they select \~10 repetitions at \~53% 1RM [@steeleAreTraineesLifting2022; @nuzzoMaximalNumberRepetitions2024]. Thus, whether self-selecting a load for a typically prescribed repetition range or self-selecting a load and self-selecting the number of repetitions to perform, most will naturally perform sets with an estimated proximity of \~10-20 repetitions shy of momentary failure.

@@

However, meta-analysis of the dose-response relationship of proximity to momentary failure for strength and hypertrophy outcomes highlights that, while there is increased hypertrophy with closer proximity to failure, there is not a clear relationship for strength [@robinsonExploringDoseResponse2024]. But a caveat is that there were limited data for proximities to failure \>10 repetitions, and that the models were adjusted for load (average loads were typically \~75-85% of one repetition maximum). Thus, the results of this meta-analysis apply to proximity to failure *after* selecting load (i.e., intraset effort) and a higher relative load requires a greater effort all else being equal. When considering the average self-selected load of \~53% 1RM coupled with a typical \~10 repetitions per set equating to proximities to failure \>10 repetitions [@steeleAreTraineesLifting2022] it seems plausible that unsupervised trainees may train with a less-than-optimal load, repetitions, and resultant effort to optimise strength outcomes. Ultimately, and in combination, supervision should provide trainees with feedback of their past performance, whether technical, effort- or program-based, and guidance toward their future performance [@fisherSupervisionStrengthTrainingthe2025].

@@

The study was conducted at a selection of strength clinics operated by Kieser Australia recruiting from the existing pool of Core members at these locations. All Kieser Australia members are prescribed the same protocol. This consists of a single set of the resistance machine exercises prescribed on their current training programme card. A load is used that should permit them to reach momentary failure within a time-under-load (TUL) of 90-120 seconds (though an upper limit of 180 seconds TUL is enforced to avoid machines being occupied for too long on the clinic floor preventing other members from using them) using a \~ 12 seconds repetition duration (i.e., \~4:4 seconds concentric:eccentric actions with a 2 isometric second hold whilst still under load with tension, not "locked-out", on the involved musculature at the end of each concentric and eccentric muscle action). Core trainees are prompted to progress load by \~5% for the next session once a TUL of \>120 seconds is achieved for a particular exercise before reaching momentary failure, and for Assisted trainees this is actioned by their supervising Exercise Scientist. Trainees are prompted to complete a session rating of perceived effort (RPE) at the end of their training session.

An acute randomised cross-over design was employed to examine participants completing an RT session with and without direct supervision with one week apart (i.e., a Core vs Assisted session). Participants were instructed to complete a single set of the resistance machine exercises on their current training programme card using their current training loads to momentary failure. The TUL performed (where a longer TUL was interpreted as indicating greater effort), and ratings of perceptions of effort (RPE) and discomfort (RPD) were recorded and compared between both conditions. This study was not pre-registered and as described below the sample size was justified based on logistical concerns and the analysis is considered exploratory.

@@

We queried the Kieser Australia database to generate random samples from historical data. We opted to generate reasonably large samples to ensure precise prior estimates, though did not use the entirety of the historical data (\~50000 members, \>10 years of data) so as to reduce computation time for Bayesian modelling. For TUL we limited data to Core and Assisted members training sessions that either were not, or were, led by an exercise scientist respectively[^3]. We took the first training session after at least 6 months of previous training at Kieser Australia had been completed by each member. Then we randomly sampled 1000 Core and 1000 Assisted members and filtered to the resistance machines used by members in the experimental study. Thus our final sample size was n = `r length(unique(prior_data_tul$id))` (Core = `r length(unique(filter(prior_data_tul, core_assisted == "core")$id))`, Assisted = `r length(unique(filter(prior_data_tul, core_assisted == "assisted")$id))`). For session RPE we performed a similar database query and randomly sampled 1000 Core and 1000 Assisted members who had reported session RPE values.

@@

All RT was performed using resistance machines (Kieser Training AG, Zurich, Switzerland) including the A1 (hip extension), A2 (torso flexion), A3 (hip abduction), B1 (leg extension), B6 (leg press), B7 (seated leg curl), C3 (torso arm i.e., pulldown), C7 (seated row), D5 (arm cross i.e., pectoral fly), D6 (chest press), D7 (seated dip), F2/F2.1 (abdominal flexion), and K2 (supported supinated grip pullup). Note, not all participants performed exercise using each machine, but had at least some of these machines in their current training cards. For the experimental study participants were instructed to attend two training sessions at least one-week apart where they completed either their current Core session (i.e., unsupervised) or an Assisted session (i.e., supervised). During the Core session the participant completed their training session as prescribed with the only difference for the Assisted session being that an exercise scientist supervised them providing instruction and motivation; however, they did not interfere in any physical way during the set (e.g. to spot or assist in completing a failing repetition). Each participant had an existing training programme card. In both conditions they utilised the current training loads for each machine that were recorded in the last session of their training programme card before participation in the study.

Participants were instructed to ensure that they continue performing the exercise to momentary failure independently of the TUL achieved for the purposes of the study i.e., if they realised during the set that their previous load selection was evidently too low for the prescribed TUL range they should continue to momentary failure regardless off the TUL acheived. Momentary failure was defined as per Steele et al. [-@steeleClarityReportingTerminology2017] i.e., the point at which, despite their greatest effort, participants are unable to continue concentrically contracting and moving the resistance, and this was clearly communicated to the participants. During each session for each exercise the trainees/trainers recorded their TUL achieved as they would do for their usual sessions using standard timers situated around the clinic in view during training specifically for this purpose (these are essentially Swiss engineered clocks with sweep movement for the second hand and positioned so that there is always one in eye line for every machine to enable recording of the TUL in seconds). These TUL recordings were then input to their training cards. In addition, they recorded their RPE and RPD in that order immediately upon completing the exercise using previously validated scales for differentiating these perceptions [@steeleDifferentiationPerceivedEffort2017]; scripts and scales are available here: [https://osf.io/ufvy8/](https://osf.io/ufvy8/).
